Component response rate variation drives stability in large complex systems

### 2018-08-10 notebook/ms.tex
Update files for PLoS Biol

\textbf{The stability of a complex system generally decreases with
increasing system size and interconnectivity, a counterintuitive result
of widespread importance across the physical, life, and social sciences.
Despite recent interest in the relationship between system properties
and stability, the effect of variation in the response rate of
individual system components remains unconsidered. Here I vary the
component response rates (\(\boldsymbol{\gamma}\)) of randomly generated
complex systems. I show that when component response rates vary, the
potential for system stability is markedly increased. Variation in
\(\boldsymbol{\gamma}\) becomes increasingly important as system size
increases, such that the largest stable complex systems would be
unstable if not for \(\boldsymbol{Var(\gamma)}\). My results reveal a
previously unconsidered driver of system stability that is likely to be
pervasive across all complex systems.}

In 1972, May (\emph{1}) first demonstrated that randomly assembled
systems of sufficient complexity are almost inevitably unstable given
infinitesimally small perturbations. Complexity in this case is defined
by the size of the system (i.e., the number of potentially interacting
components; \(S\)), its connectance (i.e., the probability that one
component will interact with another; \(C\)), and the variance of
interaction strengths (\(\sigma^{2}\)) (\emph{2}). May's finding that
the probability of local stability falls to near zero given a
sufficiently high threshold of \(\sigma\sqrt{SC}\) is broadly relevant
for understanding the dynamics and persistence of systems such as
ecological (\emph{1}--\emph{5}), neurological (\emph{6}, \emph{7}),
biochemical (\emph{8}, \emph{9}), and socio-economic
(\emph{10}--\emph{13}) networks. As such, identifying general principles
that drive stability in complex systems is of wide-ranging importance.

Randomly assembled complex systems can be represented as large square
matrices (\(\mathbf{M}\)) with \(S\) components (e.g., networks of
species (\emph{2}) or banks (\emph{11})). One element of such a matrix,
\(M_{ij}\), defines how component \(j\) affects component \(i\) in the
system at a point of equilibrium (\emph{2}). Off-diagonal elements
(\(i \neq j\)) therefore define interactions between components, while
diagonal elements (\(i = j\)) define component self-regulation (e.g.,
carrying capacity in ecological communities). Traditionally,
off-diagonal elements are assigned non-zero values with a probability
\(C\), which are sampled from a distribution with variance
\(\sigma^{2}\); diagonal elements are set to -1 (\emph{1}, \emph{2},
\emph{4}). Local system stability is assessed using eigenanalysis, with
the system being stable if the real parts of all eigenvalues
(\(\lambda\)) of \(\mathbf{M}\) are negative
(\(\max\left(\Re(\lambda)\right) < 0\)) (\emph{1}, \emph{2}). In a large
system (high \(S\)), eigenvalues are distributed uniformly (\emph{14})
within a circle centred at \(\Re = -1\) (the mean value of diagonal
elements) and \(\Im = 0\), with a radius of \(\sigma\sqrt{SC}\)
(\emph{1}, \emph{2}, \emph{4}) (Figs 1a and 2a). Local stability of
randomly assembled systems therefore becomes increasingly unlikely as
\(S\), \(C\), and \(\sigma^{2}\) increase.

May's (\emph{1}, \emph{2}) stability criterion \(\sigma\sqrt{SC} < 1\)
assumes that individual components respond to perturbations of the
system at the same rate (\(\gamma\)), but this is highly unlikely in any
complex system. In ecological communities, for example, the rate at
which population density changes following perturbation will depend on
the generation time of organisms, which might vary by orders of
magnitude among species. Species with short generation times will
respond quickly (high \(\gamma\)) to perturbations relative to species
with long generation times (low \(\gamma\)). Similarly, the speed at
which individual banks respond to perturbations in financial networks,
or individuals or institutions respond to perturbations in complex
social networks, is likely to vary. The effect of such variance on
stability has not been investigated in complex systems theory.
Intuitively, variation in \(\gamma\) might be expected to decrease
system stability by introducing a new source of variation into the
system and thereby increasing \(\sigma\). Here I show why, despite
higher \(\sigma\), complex systems in which \(\gamma\) varies are
actually more likely to be stable, especially when \(S\) is high.

@@

\textbf{Component response rates of random complex systems}. Rows in
\(\mathbf{M}\) define how a given component \(i\) is affected by other
components of the system, meaning that the rate of component response
time can be modelled by multiplying all row elements by a real scalar
value \(\gamma_{i}\) (\emph{15}). The distribution of \(\gamma\) over
\(S\) components thereby models the distribution of component response
rates. An instructive example compares one \(\mathbf{M}\) where
\(\gamma_{i} = 1\) for all \(i\) in \(S\) to the same \(\mathbf{M}\)
when half of \(\gamma_{i} = 1.95\) and half of \(\gamma_{i} = 0.05\).
This models one system in which \(\gamma\) is invariant and one in which
\(\gamma\) varies, but systems are otherwise identical (note
\(E[\gamma_{i}] = 1\) in both cases). I assume \(S = 200\),
\(C = 0.05\), and \(\sigma = 0.4\); diagonal elements are set to \(-1\)
and non-zero off-diagonal elements are drawn randomly from
\(\mathcal{N}(0, \sigma^{2})\). Rows are then multiplied by
\(\gamma_{i}\) to generate \(\mathbf{M}\). When \(\gamma_{i} = 1\),
eigenvalues of \(\mathbf{M}\) are distributed uniformly within a circle
centred at (\(-1, 0\)) with a radius of 1.265 (Fig. 1a). Hence, the real
components of eigenvalues are highly unlikely to all be negative when
all \(\gamma_{i} = 1\). But when \(\gamma_{i}\) values are separated
into two groups, eigenvalues are no longer uniformly distributed (Fig.
1b). Instead, two distinct clusters of eigenvalues appear (red circles
in Fig. 1b), one centred at (\(-1.95, 0\)) and the other centred at
(\(-0.05, 0\)). The former has a large radius, but the real components
have shifted to the left (in comparison to when \(\gamma = 1\)), while
the latter cluster has a smaller radius, but real components have
shifted to the right; all \(\Re({\lambda}) < 0\). Overall, for 1 million
randomly assembled \(\mathbf{M}\), this division between slow and fast
component response rates results in more stable systems: 1 stable given
\(\gamma = 1\) versus 32 stable given \(\gamma = \{1.95, 0.5\}\).

Higher stability in systems with variation in \(\gamma\) can be observed
by sampling \(\gamma_{i}\) values from various distributions. I focus on
a uniform distribution where \(\gamma \sim \mathcal{U}(0, 2)\) (see
Supplementary Information for other distributions, which give similar
results). As with the case of \(\gamma = \{1.95, 0.5\}\) (Fig. 1b),
\(E[\gamma] = 1\) when \(\gamma \sim \mathcal{U}(0, 2)\), allowing
comparison of \(\mathbf{M}\) before and after the addition of variation
in component response rate. Figure 2 shows a comparison of eigenvalue
distributions given \(S = 1000\), \(C = 0.05\), and \(\sigma = 0.4\). As
expected (\emph{14}), when \(\gamma = 1\), eigenvalues are distributed
uniformly in a circle centred at (\(-1, 0\)) with a radius of
\(\sigma\sqrt{SC} =\) 2.828. Uniform variation in \(\gamma\) leads to a
non-uniform distribution of eigenvalues (\emph{16}--\emph{18}), some of
which are clustered locally near the centre of the distribution, but
others of which are spread outside the former radius of 2.828 (Fig 2b).
The clustering and spreading of eigenvalues introduced by
\(Var(\gamma)\) can destabilise previously stable systems or stabilise
systems that are otherwise unstable. But where systems are otherwise too
complex to be stable given \(\gamma = 1\), the effect of \(Var(\gamma)\)
can often lead to stability above May's (\emph{1}, \emph{2}) threshold
\(\sigma\sqrt{SC} < 1\).

@@

\textbf{Targeted manipulation of \(\mathbf{\gamma}\)}. To further
investigate the potential of \(Var(\gamma)\) to be stabilising, I used a
genetic algorithm. Genetic algorithms are heuristic tools that mimic
evolution by natural selection, and are useful when the space of
potential solutions (in this case, possible combinations of \(\gamma\)
values leading to stability in a large complex system) is too large to
search exhaustively (\emph{19}). Generations of selection on \(\gamma\)
value combinations to minimise \(\max\left(\Re(\lambda)\right)\)
demonstrated the potential for \(Var(\gamma)\) to increase system
stability. Across \(S = \{2, 3, ..., 39, 40\}\), sets of \(\gamma\)
values were found that resulted in stable systems with probabilities
that were up to four orders of magnitude higher than when \(\gamma = 1\)
(Fig. 4), meaning that stability could often be achieved by manipulating
\(S\) \(\gamma\) values rather than \(S \times S\) \(\mathbf{M}\)
elements.

\textbf{System feasibility given \(\mathbf{Var(\gamma)}\)} For complex
systems in which individual system components represent the density of
some tangible quantity, it is relevant to consider the feasibility of
the system. Feasibilility assumes that the system is stable and values
of all components are positive at equilibrium (\emph{5}, \emph{20},
\emph{21}). This is of particular interest for ecological communities
because population density cannot take negative values, meaning that
ecological systems need to be feasible for stability to be biologically
realistic (\emph{20}). While my results are intended to be general to
all complex systems, and not restricted to species networks, I have also
performed a feasibility analysis on all matrices \(\mathbf{M}\) tested
for stability, and additionally for specific types of ecological
communities (\emph{2}) (e.g., competitive, mutualist, predator-prey; see
Supplementary Information). Feasibility was unaffected by
\(Var(\gamma)\), meaning that for pure interacting species networks,
variation in component response rate (i.e., species generation time)
does not affect stability at biologically realistic species densities.

@@

Here I have shown that the stability of large systems might often be
contigent upon variation in the response rates of their individual
components, meaning that factors such as rate of trait evolution (in
biological networks), transaction speed (in economic networks), or
communication speed (in social networks) need to be considered when
investigating the stability of complex systems. Variation in component
response rate becomes more likely to be critical for stability as system
size increases, and can ultimately increase the overall probability that
system stability is observed above that predicted by May's (\emph{1})
classically derived \(\sigma \sqrt{SC}\) criterion. The logic outlined
here is general, and potentially applies to any complex system in which
individual system components can vary in their reaction rates to system
perturbation.

It is important to emphasise that variation in component response rate
is not stabilising per se; that is, adding variation in component
response rate to a particular system does not necessarily increase the
probability that the system will be stable. Rather, systems that are
observed to be stable are more likely to have varying component response
rates, and for this variation to be critical to their stability (Fig.
3). This is caused by the shift to a non-uniform distribution of
eigenvalues that occurs by introducing \(Var(\gamma)\) (Fig. 1b, 2b),
which can sometimes cause all of the real components of the eigenvalues
of the community matrix \(\mathbf{M}\) to become negative, but might
also increase the real components of eigenvalues. The mathematics
underlying this shift in eigenvalue distribution has been investigated
(\emph{16}) and recently applied to questions concerning species density
and feasibility (\emph{17}, \emph{18}), but has not been interpreted as
rates of response of individual system components to perturbation.

@@

Interestingly, while complex systems were more likely to be stable given
variation in component response rate, they were not more likely to be
feasible, meaning that stability was not increased when component values
were also restricted to being positive at equilibrium. Feasibility is
important to consider, particularly for the study of ecological networks
of species (\emph{5}, \emph{18}, \emph{20}) because population densities
cannot realistically be negative. My results therefore suggest that
variation in the rate of population responses to perturbation (e.g., due
to differences in generation time among species) is unlikely to be
critical to the stability of purely multi-species interaction networks
(see also Supplementary Information). Nevertheless, ecological
interactions do not exist in isolation in empirical systems (\emph{15}),
but instead interact with evolutionary, abiotic, or social-economic
systems. The relevance of component response rate for complex system
stability should therefore not be ignored in the broader context of
ecological communities.

@@

\textbf{Component response rate variation (\(\mathbf{\gamma}\))}. In a
synthesis of eco-evolutionary feedbacks on community stability, Patel et
al. model a system that includes a vector of potentially changing
species densities (\(\mathbf{N}\)) and a vector of potentially evolving
traits (\(\mathbf{x}\))(\emph{15}). For any species \(i\) or trait
\(j\), change in species density (\(N_{i}\)) or trait value (\(x_{j}\))
with time (\(t\)) is a function of the vectors \(\mathbf{N}\) and
\(\mathbf{x}\),

@@

In the above, \(f_{i}\) and \(g_{j}\) are functions that define the
effects of all species densities and trait values on the density of a
species \(i\) and the value of trait \(j\), respectively. Patel et al.
were interested in stability when the evolution of traits was relatively
slow or fast in comparison with the change in species densities
(\emph{15}), and this is modulated in the above by the scalar
\(\epsilon\). The value of \(\epsilon\) thereby determines the timescale
separation between ecology and evolution, with high \(\epsilon\)
modelling relatively fast evolution and low \(\epsilon\) modelling
relative slow evolution (\emph{15}).

I use the same principle that Patel et al. use to modulate the relative
rate of evolution to modulate rates of component responses for \(S\)
components. Following May (\emph{1}, \emph{22}), the value of a
component \(i\) at time \(t\) (\(v_{i}(t)\)) is affected by the value of
\(j\) (\(v_{j}(t)\)) and \(j\)'s marginal effect on \(i\) (\(a_{ij}\)),
and by \(i\)'s response rate (\(\gamma_{i}\)),

@@

In matrix notation(\emph{22}),

@@

In the above, \(\mathbf{\gamma}\) is a diagonal matrix in which elements
correspond to individual component response rates. Therefore,
\(\mathbf{M} =\mathbf{\gamma} \mathbf{A}\) modulates the values of
components and can be analysed using the techniques of May (\emph{1},
\emph{16}, \emph{22}).

\textbf{Genetic algorithm}. Ideally, to investigate the potential of
\(Var(\gamma)\) for increasing the proportion of stable complex systems,
the search space of all possible \(\gamma\) vectors would be evaluated
for each unique \(\mathbf{M = \gamma A}\). This is technically
impossible because \(\gamma_{i}\) can take any real value between 0-2,
but even rounding \(\gamma_{i}\) to reasonable values would result in a
search space too large to practically explore. Under these conditions,
genetic algorithms are highly useful tools for finding practical
solutions by mimicking the process of biological evolution (\emph{19}).
In this case, the practical solution is finding vectors of
\(\mathbf{\gamma}\) that decrease the most positive real eigenvalue of
\(\mathbf{M}\). The genetic algorithm used achieves this by initialising
a large population of 1000 different potential \(\mathbf{\gamma}\)
vectors and allowing this population to evolve through a process of
mutation, crossover (swaping \(\gamma_{i}\) values between vectors),
selection, and reproduction until either a \(\mathbf{\gamma}\) vector is
found where all \(\Re(\lambda) < 0\) or some ``giving up'' critiera is
met.

@@

\textbf{System feasibility}. Dougoud et al. (\emph{20}) define the
following feasibility criteria for ecological systems characterised by
\(S\) interacting species with varying densities in a classical
Lotka-Volterra model,

@@

In the above, \(\mathbf{x^{*}}\) is the vector of species densities at
equilibrium. Feasibility is satisfied if all elements in
\(\mathbf{x^{*}}\) are positive. The matrix \(\mathbf{I}\) is the
identity matrix, and the value \(\theta\) is the strength of
intraspecific competition (diagonal elements). Diagonal values are set
to \(-1\), so \(\theta = -1\). The variable \(\delta\) is a
normalisation parameter that modulates the strength of interactions
(\(\sigma\)) for \(\mathbf{J}\). Implicitly, here \(\delta = 0\)
underlying strong interactions. Hence, \((CS)^{-\delta} = 1\), so in the
above, a diagonal matrix of -1s (\(\theta \mathbf{I}\)) is added to
\(\mathbf{J}\), which has a diagonal of all zeros and an off-diagonal
affecting species interactions (i.e., the expression \((CS)^{-\delta}\)
relates to May's(\emph{1}) stability criterion (\emph{20}) by
\(\frac{\sigma}{(CS)^{-\delta}}\sqrt{SC} < 1\), and hence for my
purposes \((CS)^{-\delta} = 1\)). Given
\(\mathbf{A} = \theta\mathbf{I + J}\), the above criteria is therefore
reduced to the below,

@@

\hypertarget{refs}{}
\hypertarget{ref-May1972}{}
1. R. M. May, Will a large complex system be stable? \emph{Nature}.
\textbf{238}, 413--414 (1972).

\hypertarget{ref-Allesina2012}{}
2. S. Allesina, S. Tang, Stability criteria for complex ecosystems.
\emph{Nature}. \textbf{483}, 205--208 (2012).

>>>>

\textbf{Key words:} Ecological networks, gene-regulatory networks,
neural networks, financial networks, system stability, random matrix
theory

\subsection{Abstract}\label{abstract}

The stability of a complex system generally decreases with increasing
system size and interconnectivity, a counterintuitive result of
widespread importance across the physical, life, and social sciences.
Despite recent interest in the relationship between system properties
and stability, the effect of variation in the response rate of
individual system components remains unconsidered. Here I vary the
component response rates (\(\boldsymbol{\gamma}\)) of randomly generated
complex systems. I show that when component response rates vary, the
potential for system stability is markedly increased. Variation in
\(\boldsymbol{\gamma}\) becomes increasingly important as system size
increases, such that the largest stable complex systems would be
unstable if not for \(\boldsymbol{Var(\gamma)}\). My results reveal a
previously unconsidered driver of system stability that is likely to be
pervasive across all complex systems.

\subsection{Author Summary}\label{author-summary}

Any set of discrete components that potentially interact can define a
complex system. Typically, properties of individual components will
cause them to respond to system perturbation at different rates, but the
consequences of such variation for system stability are unexplored. This
work makes three novel theoretical contributions: First, I show that
large complex systems with varying component response rates are more
likely to be stable than when response rates are uniform. Second, I show
that a targeted manipulation of component response rates can increase
the probability of system stability by several orders of magnitude.
Lastly, I show that varying component response rate does not affect
system feasibility. These results are general to any physical,
biological, social, or mixed system.

\subsection{Introduction}\label{introduction}

In 1972, May {[}\protect\hyperlink{ref-May1972}{1}{]} first demonstrated
that randomly assembled systems of sufficient complexity are almost
inevitably unstable given infinitesimally small perturbations.
Complexity in this case is defined by the size of the system (i.e., the
number of potentially interacting components; \(S\)), its connectance
(i.e., the probability that one component will interact with another;
\(C\)), and the variance of interaction strengths (\(\sigma^{2}\))
{[}\protect\hyperlink{ref-Allesina2012}{2}{]}. May's finding that the
probability of local stability falls to near zero given a sufficiently
high threshold of \(\sigma\sqrt{SC}\) is broadly relevant for
understanding the dynamics and persistence of systems such as ecological
{[}\protect\hyperlink{ref-May1972}{1}--\protect\hyperlink{ref-Grilli2017}{5}{]},
neurological
{[}\protect\hyperlink{ref-Gray2008}{6},\protect\hyperlink{ref-Gray2009}{7}{]},
biochemical
{[}\protect\hyperlink{ref-Rosenfeld2009}{8},\protect\hyperlink{ref-MacArthur2010}{9}{]},
and socio-economic
{[}\protect\hyperlink{ref-May2008}{10}--\protect\hyperlink{ref-Bardoscia2017}{13}{]}
networks. As such, identifying general principles that drive stability
in complex systems is of wide-ranging importance.

Randomly assembled complex systems can be represented as large square
matrices (\(\mathbf{M}\)) with \(S\) components (e.g., networks of
species {[}\protect\hyperlink{ref-Allesina2012}{2}{]} or banks
{[}\protect\hyperlink{ref-Haldane2011}{11}{]}). One element of such a
matrix, \(M_{ij}\), defines how component \(j\) affects component \(i\)
in the system at a point of equilibrium
{[}\protect\hyperlink{ref-Allesina2012}{2}{]}. Off-diagonal elements
(\(i \neq j\)) therefore define interactions between components, while
diagonal elements (\(i = j\)) define component self-regulation (e.g.,
carrying capacity in ecological communities). Traditionally,
off-diagonal elements are assigned non-zero values with a probability
\(C\), which are sampled from a distribution with variance
\(\sigma^{2}\); diagonal elements are set to -1
{[}\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2015}{4}{]}.
Local system stability is assessed using eigenanalysis, with the system
being stable if the real parts of all eigenvalues (\(\lambda\)) of
\(\mathbf{M}\) are negative (\(\max\left(\Re(\lambda)\right) < 0\))
{[}\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}{]}.
In a large system (high \(S\)), eigenvalues are distributed uniformly
{[}\protect\hyperlink{ref-Tao2010}{14}{]} within a circle centred at
\(\Re = -1\) (the mean value of diagonal elements) and \(\Im = 0\), with
a radius of \(\sigma\sqrt{SC}\)
{[}\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2015}{4}{]}
(Figs 1a and 2a). Local stability of randomly assembled systems
therefore becomes increasingly unlikely as \(S\), \(C\), and
\(\sigma^{2}\) increase.

May's
{[}\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}{]}
stability criterion \(\sigma\sqrt{SC} < 1\) assumes that individual
components respond to perturbations of the system at the same rate
(\(\gamma\)), but this is highly unlikely in any complex system. In
ecological communities, for example, the rate at which population
density changes following perturbation will depend on the generation
time of organisms, which might vary by orders of magnitude among
species. Species with short generation times will respond quickly (high
\(\gamma\)) to perturbations relative to species with long generation
times (low \(\gamma\)). Similarly, the speed at which individual banks
respond to perturbations in financial networks, or individuals or
institutions respond to perturbations in complex social networks, is
likely to vary. The effect of such variance on stability has not been
investigated in complex systems theory. Intuitively, variation in
\(\gamma\) might be expected to decrease system stability by introducing
a new source of variation into the system and thereby increasing
\(\sigma\). Here I show why, despite higher \(\sigma\), complex systems
in which \(\gamma\) varies are actually more likely to be stable,
especially when \(S\) is high.

@@

\textbf{Component response rates of random complex systems}. Rows in
\(\mathbf{M}\) define how a given component \(i\) is affected by other
components of the system, meaning that the rate of component response
time can be modelled by multiplying all row elements by a real scalar
value \(\gamma_{i}\) {[}\protect\hyperlink{ref-Patel2018}{15}{]}. The
distribution of \(\gamma\) over \(S\) components thereby models the
distribution of component response rates. An instructive example
compares one \(\mathbf{M}\) where \(\gamma_{i} = 1\) for all \(i\) in
\(S\) to the same \(\mathbf{M}\) when half of \(\gamma_{i} = 1.95\) and
half of \(\gamma_{i} = 0.05\). This models one system in which
\(\gamma\) is invariant and one in which \(\gamma\) varies, but systems
are otherwise identical (note \(E[\gamma_{i}] = 1\) in both cases). I
assume \(S = 200\), \(C = 0.05\), and \(\sigma = 0.4\); diagonal
elements are set to \(-1\) and non-zero off-diagonal elements are drawn
randomly from \(\mathcal{N}(0, \sigma^{2})\). Rows are then multiplied
by \(\gamma_{i}\) to generate \(\mathbf{M}\). When \(\gamma_{i} = 1\),
eigenvalues of \(\mathbf{M}\) are distributed uniformly within a circle
centred at (\(-1, 0\)) with a radius of 1.265 (Fig. 1a). Hence, the real
components of eigenvalues are highly unlikely to all be negative when
all \(\gamma_{i} = 1\). But when \(\gamma_{i}\) values are separated
into two groups, eigenvalues are no longer uniformly distributed (Fig.
1b). Instead, two distinct clusters of eigenvalues appear (red circles
in Fig. 1b), one centred at (\(-1.95, 0\)) and the other centred at
(\(-0.05, 0\)). The former has a large radius, but the real components
have shifted to the left (in comparison to when \(\gamma = 1\)), while
the latter cluster has a smaller radius, but real components have
shifted to the right; all \(\Re({\lambda}) < 0\). Overall, for 1 million
randomly assembled \(\mathbf{M}\), this division between slow and fast
component response rates results in more stable systems: 1 stable given
\(\gamma = 1\) versus 32 stable given \(\gamma = \{1.95, 0.5\}\).

Higher stability in systems with variation in \(\gamma\) can be observed
by sampling \(\gamma_{i}\) values from various distributions. I focus on
a uniform distribution where \(\gamma \sim \mathcal{U}(0, 2)\) (see
Supplementary Information for other distributions, which give similar
results). As with the case of \(\gamma = \{1.95, 0.5\}\) (Fig. 1b),
\(E[\gamma] = 1\) when \(\gamma \sim \mathcal{U}(0, 2)\), allowing
comparison of \(\mathbf{M}\) before and after the addition of variation
in component response rate. Figure 2 shows a comparison of eigenvalue
distributions given \(S = 1000\), \(C = 0.05\), and \(\sigma = 0.4\). As
expected {[}\protect\hyperlink{ref-Tao2010}{14}{]}, when \(\gamma = 1\),
eigenvalues are distributed uniformly in a circle centred at (\(-1, 0\))
with a radius of \(\sigma\sqrt{SC} =\) 2.828. Uniform variation in
\(\gamma\) leads to a non-uniform distribution of eigenvalues
{[}\protect\hyperlink{ref-Ahmadian2015}{16}--\protect\hyperlink{ref-Stone2017}{18}{]},
some of which are clustered locally near the centre of the distribution,
but others of which are spread outside the former radius of 2.828 (Fig
2b). The clustering and spreading of eigenvalues introduced by
\(Var(\gamma)\) can destabilise previously stable systems or stabilise
systems that are otherwise unstable. But where systems are otherwise too
complex to be stable given \(\gamma = 1\), the effect of \(Var(\gamma)\)
can often lead to stability above May's
{[}\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}{]}
threshold \(\sigma\sqrt{SC} < 1\).

@@

\textbf{Targeted manipulation of \(\mathbf{\gamma}\)}. To further
investigate the potential of \(Var(\gamma)\) to be stabilising, I used a
genetic algorithm. Genetic algorithms are heuristic tools that mimic
evolution by natural selection, and are useful when the space of
potential solutions (in this case, possible combinations of \(\gamma\)
values leading to stability in a large complex system) is too large to
search exhaustively {[}\protect\hyperlink{ref-Hamblin2013}{19}{]}.
Generations of selection on \(\gamma\) value combinations to minimise
\(\max\left(\Re(\lambda)\right)\) demonstrated the potential for
\(Var(\gamma)\) to increase system stability. Across
\(S = \{2, 3, ..., 39, 40\}\), sets of \(\gamma\) values were found that
resulted in stable systems with probabilities that were up to four
orders of magnitude higher than when \(\gamma = 1\) (Fig. 4), meaning
that stability could often be achieved by manipulating \(S\) \(\gamma\)
values rather than \(S \times S\) \(\mathbf{M}\) elements.

\textbf{System feasibility given \(\mathbf{Var(\gamma)}\)} For complex
systems in which individual system components represent the density of
some tangible quantity, it is relevant to consider the feasibility of
the system. Feasibilility assumes that the system is stable and values
of all components are positive at equilibrium
{[}\protect\hyperlink{ref-Grilli2017}{5},\protect\hyperlink{ref-Dougoud2018}{20},\protect\hyperlink{ref-Song2018}{21}{]}.
This is of particular interest for ecological communities because
population density cannot take negative values, meaning that ecological
systems need to be feasible for stability to be biologically realistic
{[}\protect\hyperlink{ref-Dougoud2018}{20}{]}. While my results are
intended to be general to all complex systems, and not restricted to
species networks, I have also performed a feasibility analysis on all
matrices \(\mathbf{M}\) tested for stability, and additionally for
specific types of ecological communities
{[}\protect\hyperlink{ref-Allesina2012}{2}{]} (e.g., competitive,
mutualist, predator-prey; see Supplementary Information). Feasibility
was unaffected by \(Var(\gamma)\), meaning that for pure interacting
species networks, variation in component response rate (i.e., species
generation time) does not affect stability at biologically realistic
species densities.

@@

Here I have shown that the stability of large systems might often be
contigent upon variation in the response rates of their individual
components, meaning that factors such as rate of trait evolution (in
biological networks), transaction speed (in economic networks), or
communication speed (in social networks) need to be considered when
investigating the stability of complex systems. Variation in component
response rate becomes more likely to be critical for stability as system
size increases, and can ultimately increase the overall probability that
system stability is observed above that predicted by May's
{[}\protect\hyperlink{ref-May1972}{1}{]} classically derived
\(\sigma \sqrt{SC}\) criterion. The logic outlined here is general, and
potentially applies to any complex system in which individual system
components can vary in their reaction rates to system perturbation.

It is important to emphasise that variation in component response rate
is not stabilising per se; that is, adding variation in component
response rate to a particular system does not necessarily increase the
probability that the system will be stable. Rather, systems that are
observed to be stable are more likely to have varying component response
rates, and for this variation to be critical to their stability (Fig.
3). This is caused by the shift to a non-uniform distribution of
eigenvalues that occurs by introducing \(Var(\gamma)\) (Fig. 1b, 2b),
which can sometimes cause all of the real components of the eigenvalues
of the community matrix \(\mathbf{M}\) to become negative, but might
also increase the real components of eigenvalues. The mathematics
underlying this shift in eigenvalue distribution has been investigated
{[}\protect\hyperlink{ref-Ahmadian2015}{16}{]} and recently applied to
questions concerning species density and feasibility
{[}\protect\hyperlink{ref-Gibbs2017}{17},\protect\hyperlink{ref-Stone2017}{18}{]},
but has not been interpreted as rates of response of individual system
components to perturbation.

@@

Interestingly, while complex systems were more likely to be stable given
variation in component response rate, they were not more likely to be
feasible, meaning that stability was not increased when component values
were also restricted to being positive at equilibrium. Feasibility is
important to consider, particularly for the study of ecological networks
of species
{[}\protect\hyperlink{ref-Grilli2017}{5},\protect\hyperlink{ref-Stone2017}{18},\protect\hyperlink{ref-Dougoud2018}{20}{]}
because population densities cannot realistically be negative. My
results therefore suggest that variation in the rate of population
responses to perturbation (e.g., due to differences in generation time
among species) is unlikely to be critical to the stability of purely
multi-species interaction networks (see also Supplementary Information).
Nevertheless, ecological interactions do not exist in isolation in
empirical systems {[}\protect\hyperlink{ref-Patel2018}{15}{]}, but
instead interact with evolutionary, abiotic, or social-economic systems.
The relevance of component response rate for complex system stability
should therefore not be ignored in the broader context of ecological
communities.

@@

\textbf{Component response rate variation (\(\mathbf{\gamma}\))}. In a
synthesis of eco-evolutionary feedbacks on community stability, Patel et
al. model a system that includes a vector of potentially changing
species densities (\(\mathbf{N}\)) and a vector of potentially evolving
traits (\(\mathbf{x}\)){[}\protect\hyperlink{ref-Patel2018}{15}{]}. For
any species \(i\) or trait \(j\), change in species density (\(N_{i}\))
or trait value (\(x_{j}\)) with time (\(t\)) is a function of the
vectors \(\mathbf{N}\) and \(\mathbf{x}\),

@@

In the above, \(f_{i}\) and \(g_{j}\) are functions that define the
effects of all species densities and trait values on the density of a
species \(i\) and the value of trait \(j\), respectively. Patel et al.
were interested in stability when the evolution of traits was relatively
slow or fast in comparison with the change in species densities
{[}\protect\hyperlink{ref-Patel2018}{15}{]}, and this is modulated in
the above by the scalar \(\epsilon\). The value of \(\epsilon\) thereby
determines the timescale separation between ecology and evolution, with
high \(\epsilon\) modelling relatively fast evolution and low
\(\epsilon\) modelling relative slow evolution
{[}\protect\hyperlink{ref-Patel2018}{15}{]}.

I use the same principle that Patel et al. use to modulate the relative
rate of evolution to modulate rates of component responses for \(S\)
components. Following May
{[}\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-May1973}{22}{]},
the value of a component \(i\) at time \(t\) (\(v_{i}(t)\)) is affected
by the value of \(j\) (\(v_{j}(t)\)) and \(j\)'s marginal effect on
\(i\) (\(a_{ij}\)), and by \(i\)'s response rate (\(\gamma_{i}\)),

@@

In matrix notation{[}\protect\hyperlink{ref-May1973}{22}{]},

@@

In the above, \(\mathbf{\gamma}\) is a diagonal matrix in which elements
correspond to individual component response rates. Therefore,
\(\mathbf{M} =\mathbf{\gamma} \mathbf{A}\) modulates the values of
components and can be analysed using the techniques of May
{[}\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Ahmadian2015}{16},\protect\hyperlink{ref-May1973}{22}{]}.

\textbf{Genetic algorithm}. Ideally, to investigate the potential of
\(Var(\gamma)\) for increasing the proportion of stable complex systems,
the search space of all possible \(\gamma\) vectors would be evaluated
for each unique \(\mathbf{M = \gamma A}\). This is technically
impossible because \(\gamma_{i}\) can take any real value between 0-2,
but even rounding \(\gamma_{i}\) to reasonable values would result in a
search space too large to practically explore. Under these conditions,
genetic algorithms are highly useful tools for finding practical
solutions by mimicking the process of biological evolution
{[}\protect\hyperlink{ref-Hamblin2013}{19}{]}. In this case, the
practical solution is finding vectors of \(\mathbf{\gamma}\) that
decrease the most positive real eigenvalue of \(\mathbf{M}\). The
genetic algorithm used achieves this by initialising a large population
of 1000 different potential \(\mathbf{\gamma}\) vectors and allowing
this population to evolve through a process of mutation, crossover
(swaping \(\gamma_{i}\) values between vectors), selection, and
reproduction until either a \(\mathbf{\gamma}\) vector is found where
all \(\Re(\lambda) < 0\) or some ``giving up'' critiera is met.

@@

\textbf{System feasibility}. Dougoud et al.
{[}\protect\hyperlink{ref-Dougoud2018}{20}{]} define the following
feasibility criteria for ecological systems characterised by \(S\)
interacting species with varying densities in a classical Lotka-Volterra
model,

@@

In the above, \(\mathbf{x^{*}}\) is the vector of species densities at
equilibrium. Feasibility is satisfied if all elements in
\(\mathbf{x^{*}}\) are positive. The matrix \(\mathbf{I}\) is the
identity matrix, and the value \(\theta\) is the strength of
intraspecific competition (diagonal elements). Diagonal values are set
to \(-1\), so \(\theta = -1\). The variable \(\delta\) is a
normalisation parameter that modulates the strength of interactions
(\(\sigma\)) for \(\mathbf{J}\). Implicitly, here \(\delta = 0\)
underlying strong interactions. Hence, \((CS)^{-\delta} = 1\), so in the
above, a diagonal matrix of -1s (\(\theta \mathbf{I}\)) is added to
\(\mathbf{J}\), which has a diagonal of all zeros and an off-diagonal
affecting species interactions (i.e., the expression \((CS)^{-\delta}\)
relates to May's{[}\protect\hyperlink{ref-May1972}{1}{]} stability
criterion {[}\protect\hyperlink{ref-Dougoud2018}{20}{]} by
\(\frac{\sigma}{(CS)^{-\delta}}\sqrt{SC} < 1\), and hence for my
purposes \((CS)^{-\delta} = 1\)). Given
\(\mathbf{A} = \theta\mathbf{I + J}\), the above criteria is therefore
reduced to the below,

@@

\hypertarget{refs}{}
\hypertarget{ref-May1972}{}
1. May RM. Will a large complex system be stable? Nature. 1972;238:
413--414.

\hypertarget{ref-Allesina2012}{}
2. Allesina S, Tang S. Stability criteria for complex ecosystems.
Nature. Nature Publishing Group; 2012;483: 205--208.
doi:\href{https://doi.org/10.1038/nature10832}{10.1038/nature10832}
